Future challenges for Research Software Engineering

### 2025-09-25 future_challenges_for_RSEng.tex
MHagdorn -  yet another read through (#71)
* feat(#44): fix pronouns

* feat(#44): fix tense

* feat(#44): fix tense

* split sentence

* swap words

* split long sentence

* replace word

* drop paragraph as it is already a few lines above

* reduce the number of certains

* fixes(#56): sort out incomplete sentence

* fixes(#56): rename goal boxes

* fixes(#56): swap sentences around to make it clearer that now is 2025

* minor fix

* less hope

* be more assertive

* make it clearer that the workshop discussions are meant

* reword and remove redundancy

* insert word

* insert comma

* drop

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software
was only published a couple of years ago~\cite{Anzt2021},
while the foundational competencies of an RSE have only recently been defined in detail~\cite{Goth2024}.
Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
Here, we present the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the German RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.
Since this paper focuses on the outputs of the deRSE25 conference, we present our thoughts and perspectives in a largely German-focused context.
At the same time, we recognise, the wider international growth of RSEng as a very important development
for people undertaking software development within the research community.
Much of our work and our discussions are also building on wider perspectives for the future of RSEng that are currently under
investigation among RSEng communities in other countries.
While taking a German-focused view in this paper, we therefore aim to highlight resources and materials from
the wider international community where we feel this is relevant.

@@

\begin{story}{The Definition of RSEs}{defrse}
Kim and Kay chuckle when they look at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
When labs falter under algorithmic bias or unclear model provenance, it is Kim and Kay who transform the chaos back into meaningful order.
They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
Their job titles have changed, but their role as bridges between logic and life remains the same.
In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.

@@

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious.
Kim is currently designing the software service stack at the university.
She increasingly struggles to find compelling Open Source alternatives that enables them to keep complete control of their data
rather than relying on software and tools that are backed by large corporate entities.
So far, she still has the support of the university board, since it values strategic autonomy.
Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
There, she gained valuable experience working in an industrial setting.
However, she was very happy when an opportunity arose to work in a central
RSE department with an open-ended contract. Apart from her main responsibility
of developing and maintaining software for an affordable water monitoring system
used in the global south, she now also teaches undergraduates digital literacy.
\end{story}

\subsection{Transformations in Academic Institutions}
\begin{whatis}{}{academic}
Research software is a central part of scientific practice, on the same level as data and publications.
This is supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
Some central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
Specialised clusters have formed, which pool the skills of specialised RSEs with RSE units specialising in certain tasks like data management, HPC, or visualisation.
At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
Additionally, researchers interact seamlessly with infrastructures for using research software.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of generations.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
Some specialisation trends have already been observed in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated},
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

>>>>

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software has only been raised by the RSE community a couple of years ago~\cite{Anzt2021},
while the foundational competencies of an RSE have just recently been defined in detail~\cite{Goth2024}.
Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design~\cite{Dehne2025-draft}.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are standard practice in software engineering (SE)~\cite{Katz2023,Khan2021,Hu2023,Bosch2016a,Boehm2011},
so far future projections are far less common for the RSEng domain.
We present here the results of a workshop discussion (notes content is archived here \cite{goth2025_WSPad}) at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}?
Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}},
without claiming to represent the entire deRSE community.
The material presented here is based on the output of the workshop participants, but extended with the authors' ideas.
Since deRSE25 was co-located with SE25, the annual conference for the Software Engineering community in Germany, the discussion was also open to software engineers,
but no list or profile of the participants is available.
The workshop hosts felt, that the majority of participants were from the RSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how RSEng should, or could, develop.

@@

\begin{story}{The Definition of RSEs}{defrse}
Kim and Kay chuckled when they looked at their original job descriptions from 2020: ``build tools for scientists, manage data, write reusable code.''
Now, their roles are unrecognisable: half AI whisperer, half research ethicist, and part-time diplomat negotiating between synthetic and human collaborators.
They now not only manage software, but also the personalities of autonomous lab agents and negotiate data-sharing agreements with international AI consortia.
Despite the tech upheaval, their deep knowledge of scientific workflows and the messy, human way science actually gets done remains indispensable.
When labs faltered under algorithmic bias or unclear model provenance, it was Kim and Kay who transformed the chaos back into meaningful order.
They understand that science does not move in straight lines, and that trust between collaborators, institutions, and even machines has to be built, not coded.
Their job titles have changed, but their role as bridges between logic and life remains the same.
In a future of shifting code and constant reinvention, it is their understanding of people that never went out of date.

@@

A code of ethics and professional conduct for RSEs should be developed by institutions together with RSEs themselves.
This can then form the basis for institutional policies and guidelines as well as serve as a background for individual developers to consider how software can be used.

\begin{story}{The greater world of Kim and Kay}{ethics}
10 years have passed since Kim and Kay started on their journey of being RSEs. Back then,
at the universities of Eden and Utopia, life was more orderly and harmonious.
Kim is currently designing the software service stack at the university.
She increasingly struggles to find compelling Open Source alternatives that enables them to keep complete control of their data
rather than relying on software and tools that are backed by large corporate entities.
So far, she still has the support of the university board, since it values strategic autonomy.
Kay had left academia to work in an industrial R\&D department because of a more reliable employment.
There he gained valuable experience working in an industrial setting.
However, he was very happy when an opportunity arose to work in a central
RSE department with an open-ended contract. Apart from his main responsibility
of developing and maintaining software for an affordable water monitoring system
used in the global south, he now also teaches undergraduates digital literacy.
\end{story}

\subsection{Transformations in Academic Institutions}
\begin{whatis}{}{academic}
Research software is a central part of scientific practice, on the same level as data and publications.
This will be supported by the first generations of RSE Master's graduates that find their way into research departments, and accelerate the pace of science.
Similar to RSEs, who are widely known and understood by university boards in 2035, academic staff members at any level are aware of the concept of RSE units.
Universities actively use the presence of an RSE Department as a means to differentiate themselves from other universities.
Some central RSE units will have local RSE representatives, or even small RSE groups that handle tasks within a department, overseeing specific projects~\cite{Kempf2025-draft}.
Specialised clusters have formed, which pool the skills of certain specialised RSEs with certain RSE units specialising in certain tasks like data management, HPC, or visualisation.
At universities which create such RSE units, it can be observed that researchers naturally accept them as part of their available services.
Additionally, researchers will seamlessly interact with infrastructures for using research software.
Replacement of staff reaching retirement, combined with the intensified research interest in RSE, has led to the creation of more RSE professorships.
\end{whatis}
Academia, while claiming to be innovative~\cite{wzvgbayern},
has organisationally often adopted time scales for change that are more reminiscent
to institutions thinking on the time scale of societal generations.
The RSE community should plan ahead and use opportunities on that horizon.
If in the time-frame of the next decade, a large number of permanent staff reach retirement and
can be replaced by a new generation of researchers and technical professionals who have seen the
digital transformation first-hand and maybe have actively participated in it through various grassroots initiatives.
This generational change opens up the opportunity for an active change within the research landscape.
A simple possibility would be to realign newly appointed professorships with a focus towards software development within their domain.
A second more progressive step would be to actually appoint new cross-faculty professorships.
On the other hand, the presence, even today, of initiatives like \emph{the Hidden Ref}~\cite{hiddenref}
hints at the potential for specialisation of research tasks.
This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
Some specialisation trends have already been observed in recent years,
such as the HPC-RSE and AI-RSE, which are quite recent developments,
and whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
We foresee new RSE specialisations emerging, \eg, in the fields of quantum computing,
such as the Quantum-RSE, or in digital research infrastructure.
We note the SE community has also independently expressed interest in creating an \emph{infrastructure engineering} role~\cite{Sochat2024Infra}
and funding bodies seem willing to promote such career pathways~\cite{UKRI2025Infra,EuroHPC2024Federated}.
Taking these trends together the generational change should be used as an opportunity to reshape organisational structures to better reflect the work people in research
are actually doing. The RSE community should take an active role in shaping that discussion in this direction till 2035.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum
